A tractable model for D-optimal design in infinite dimensional Bayesian inverse problems.

### 2020-07-20 paper.tex
Could it be? Is it done???

\section{Analysis of Optimal Designs --- Vanishing Model Error}
When $\modcov = 0$, then $\Sigma= \sigma^2I$. The necessary
first-order condition for optimality \eqref{eq:conditions} becomes
\begin{equation}\label{eq:eigenproblem}
  \sigma^{-2}\fwd \postcov \fwd^* \obs^* = \obs^* \Xi. 
\end{equation}  
This is an eigenvalue problem for the self-adjoint operator
$\sigma^{-2}\fwd \postcov \fwd^*$ where the rows of $\obs$, namely
$\meas_j,j=1,...m$, are the eigenvectors. If $\fwd$ is invertible,
then we may write:
\begin{align}\label{eq:mod conditions}
  \begin{split}
  \obs^* \Xi &= \sigma^{-2}\fwd \postcov \fwd^* \obs^*  \\
  %
  %
  %
  &= \sigma^{-2} \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* \obs^* \\
  %
  %
  %
  &= \sigma^{-2} \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1} \obs^*,
  \end{split}
\end{align}  
which presents the optimal design problem as a problem on $\ban$ and
not on $\hil$ --- the term $(\fwd \prcov \fwd^*)^{-1}$ is the prior
precision in $\ban$ and the big parentheses enclose the posterior
precision in $\ban$. This expression makes sense even if $\fwd$ is not
invertible --- consider $\fwd: \ker \fwd^{\perp} \to \text{Im} \fwd
\subseteq \ban$ and this does not change the optimal design problem at
all, since we will never take measurements that have a component in
$\fwd^{\perp}$ (this is intuitive by Occam's razor but can also be
understood from Corollary \ref{cor:zero mod err}).

@@

      \addlegendentry{prior}
      \addlegendentry{$a_1$}
      \addlegendentry{$a_2$}
      \addlegendentry{$a_3$}
      \addlegendentry{$a_4$}
      \addlegendentry{$a_5$}
      \addlegendentry{$a_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Posterior precision per eigenvector. Each measurement
    $\func_i$ increases precision uniformly across the lowest precision
    eigenvectors. We see that $\func_2$ and $\func_3$ are repeated
    measurements, as well as $\func_5$ and $\func_6$.}
  \label{fig:clusterization}
\end{figure}
with the norm constraints on $\meas_j,j=1,...,m$. We show in the
appedix (Lemma \ref{lemma:free}) we can find $\meas_1,...,\meas_m$
such that $\{\eta_i\}_{i=1}^k$ have any value we desire, subject only
to the restriction that we do not change the trace of $\obs^*\obs$, so
that $\sum_{i=1}^k \eta_i = m$. By concavity of $\log$, the highest
increase in the design criterion is gained by increasing
$\lambda_i^{-1} + \sigma^{-2}\eta_i$ where it is small. So, no weight
should be given to (i.e. measurement taken of) the $k$th eigenvector
before $\lambda_i^{-1} + \sigma^{-2}\eta_i \geq \lambda_k^{-1} \forall
i \neq k$. Thus, we should choose $\{\eta_i\}_{i=1}^k$ as follows. We
would increase $\lambda_1^{-1} + \sigma^{-2} \eta_1$ until it equals
$\lambda_2^{-1}$. Then, $\lambda_i^{-1} + \sigma^{-2} \eta_i,i=1,2$ is
increased until $\lambda_i^{-1} + \sigma^{-2} \eta_i =
\lambda_3^{-1},i=1,2$ and so forth.

@@

\section{Conclusion}\label{section:conclusion}
Our relaxed model gives us insight to D-optimal designs. We see that
in an ideal settings, we should not reduce uncertainty in any
direction (eigenvector) before it is the largest uncertainty present.

%% We believe the model presented is not too far from reality. After all,
%% taking any measurement (point evaluation or other) is just recording
%% the value of a linear combination of individual eigenvectors of the
%% prior. In light of this, we believe the reason for clusterization is close
%% to the argument made in section \ref{subsub:first attempt}. The reason
%% we observe clusterization in real life problems is probably due to
%% symmetry breaking, although we cannot show that rigorously.

An important consequence of the argument in \ref{section:vanishing} is
that we can find an approximately optimal design by means of linear
algebra. Here is how this should be done.
\begin{enumerate}

   \item Find eigenvalues $\lambda_1 \geq \lambda_2 \geq ...$ and
     corresponding eigenvectors $\ev_i$ of $\fwd \prcov \fwd^*$.
   \item Denote the (assumed known) variance of the measurement noise
     $\sigma^2$.
   \item Decide how many measurements you can take and denote that
     number $m$. This is the total precision one can allocate to the
     different eigenvectors. Next, we see how to allocate these.
    \item Find $k$ and $\{\eta_i\}_{i=1}^k$, $\eta_i > 0, i=1,...,k$
      such that:
      \begin{itemize}
        \item $\sum_{i=1}^k \eta_i = m$ --- the total measurement precision available.
      \item $\lambda_i^{-1} + \sigma^{-2}\eta_i = \lambda_k^{-1} +
        \sigma^{-2}\eta_k$ for $i=1,...,k-1$ --- the posterior
        precision is equal for all $k$ eigenvectors.
      \item $\lambda_{k+1}^{-1} \geq \lambda_k^{-1} + \eta_k$ --- no
        information is gained for the rest of the eigenvectors.
      \end{itemize}
      \item Use the construction in Lemma \ref{lemma:free} to find the
        optimal measurements in terms of the eigenvectors $\ev_i$ ---
        without \emph{any} optimization.
      \item Find point measurements that approximate those in the domain. 
\end{enumerate}

\section{Acknowledgements}
This study was conducted during when the author was a PhD student of
Prof. Georg Stadler in New York University's Courant Institute.

>>>>

\section{Analysis of Optimal Designs --- Vanishing Model Error}\label{section:vanishing}
When $\modcov = 0$, then $\Sigma= \sigma^2I$. The necessary
first-order condition for optimality \eqref{eq:conditions} becomes
\begin{equation}\label{eq:eigenproblem}
  \sigma^{-2}\fwd \postcov \fwd^* \obs^* = \obs^* \Xi. 
\end{equation}  
This is an eigenvalue problem for the self-adjoint operator
$\sigma^{-2}\fwd \postcov \fwd^*$ where the rows of $\obs$, namely
$\meas_j,j=1,...m$, are the eigenvectors. If $\fwd$ is invertible,
then we may write:
\begin{align}\label{eq:mod conditions}
  \begin{split}
  \obs^* \Xi &= \sigma^{-2}\fwd \postcov \fwd^* \obs^*  \\
  %
  %
  %
  &= \sigma^{-2} \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* \obs^* \\
  %
  %
  %
  &= \sigma^{-2} \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1} \obs^*,
  \end{split}
\end{align}  
which presents the optimal design problem as a problem on $\ban$ and
not on $\hil$ --- the term $(\fwd \prcov \fwd^*)^{-1}$ is the prior
precision in $\ban$ and the big parentheses enclose the posterior
precision in $\ban$. This expression makes sense even if $\fwd$ is not
invertible --- consider $\fwd: \ker \fwd^{\perp} \to \text{Im} \fwd
\subseteq \ban$ and this does not change the optimal design problem at
all, since we will never take measurements that have a component in
$\fwd^{\perp}$ (this is intuitive by Occam's razor but can also be
understood from Corollary \ref{cor:zero mod err}).

@@

      \addlegendentry{prior}
      \addlegendentry{$a_1$}
      \addlegendentry{$a_2$}
      \addlegendentry{$a_3$}
      \addlegendentry{$a_4$}
      \addlegendentry{$a_5$}
      \addlegendentry{$a_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Posterior precision per eigenvector. Each measurement
    icreases precision uniformly across the lowest precision
    eigenvectors. We see that $\func_2$ and $\func_3$ are repeated
    measurements, as well as $\func_5$ and $\func_6$.}
  \label{fig:clusterization}
\end{figure}
with the norm constraints on $\meas_j,j=1,...,m$. We show in the
appedix (Lemma \ref{lemma:free}) we can find $\meas_1,...,\meas_m$
such that $\{\eta_i\}_{i=1}^k$ have any value we desire, subject only
to the restriction that we do not change the trace of $\obs^*\obs$, so
that $\sum_{i=1}^k \eta_i = m$. By concavity of $\log$, the highest
increase in the design criterion is gained by increasing
$\lambda_i^{-1} + \sigma^{-2}\eta_i$ where it is small. So, no weight
should be given to (i.e. measurement taken of) the $k$th eigenvector
before $\lambda_i^{-1} + \sigma^{-2}\eta_i \geq \lambda_k^{-1} \forall
i \neq k$. Thus, we should choose $\{\eta_i\}_{i=1}^k$ as follows. We
would increase $\lambda_1^{-1} + \sigma^{-2} \eta_1$ until it equals
$\lambda_2^{-1}$. Then, $\lambda_i^{-1} + \sigma^{-2} \eta_i,i=1,2$ is
increased until $\lambda_i^{-1} + \sigma^{-2} \eta_i =
\lambda_3^{-1},i=1,2$ and so forth.

@@

\section{Conclusion}\label{section:conclusion}
Our relaxed model gives us insight to D-optimal designs. We see that
in an ideal settings, we should not reduce uncertainty in any
direction (eigenvector) before it is the largest uncertainty present.
We show how and why repeated measurements can give rise to a D-optimal
design. There is more work to be done in understanding exactly what
causes such designs to be strictly better than others.
%% We believe the model presented is not too far from reality. After all,
%% taking any measurement (point evaluation or other) is just recording
%% the value of a linear combination of individual eigenvectors of the
%% prior. In light of this, we believe the reason for clusterization is close
%% to the argument made in section \ref{subsub:first attempt}. The reason
%% we observe clusterization in real life problems is probably due to
%% symmetry breaking, although we cannot show that rigorously.

A consequence of the argument in section \ref{section:vanishing} is
that we can find an approximately optimal design by means of linear
algebra. Here is how this should be done.
\begin{enumerate}

   \item Find eigenvalues $\lambda_1 \geq \lambda_2 \geq ...$ and
     corresponding eigenvectors $\ev_i$ of $\fwd \prcov \fwd^*$.
   \item Denote the (assumed known) variance of the measurement noise
     $\sigma^2$.
   \item Decide how many measurements can be taken and denote that
     number $m$. This is the total precision that can be allocated to
     the different eigenvectors.
    \item Find $k$ and $\{\eta_i\}_{i=1}^k$, $\eta_i > 0, i=1,...,k$
      such that:
      \begin{itemize}
        \item $\sum_{i=1}^k \eta_i = m$ --- the total measurement precision available.
      \item $\lambda_i^{-1} + \sigma^{-2}\eta_i = \lambda_k^{-1} +
        \sigma^{-2}\eta_k$ for $i=1,...,k-1$ --- the posterior
        precision is equal for all $k$ eigenvectors.
      \item $\lambda_{k+1}^{-1} \geq \lambda_k^{-1} + \eta_k$ --- no
        information is gained for the rest of the eigenvectors.
      \end{itemize}
      \item Use the construction in Lemma \ref{lemma:free} to find the
        optimal measurements in terms of the eigenvectors $\ev_i$ ---
        without \emph{any} optimization.
      \item Find point measurements that approximate those in the domain. 
\end{enumerate}

\section{Acknowledgements}
This study was conducted during when I was a PhD student of
Prof. Georg Stadler in New York University's Courant Institute. I
would like to thank him for his great mentorship.
